A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-14 Manuscript/PNAS-Adams-02.Rmd
vsn 2

output: rticles::pnas_article
---
Investigating macroevolutionary patterns of trait variation requires a phylogenetic perspective, because the shared ancestry among species violates the assumption of independence among trait values that is common for statistical tests [@Felsenstein1985; @HarveyPagel1991]. Accounting for this evolutionary non-independence is the purview of *phylogenetic comparative methods* (PCMs): a suite of analytical tools that condition trends in the data on the phylogenetic relatedness of observations [@Grafen1989; @GarlandIves2000; @Rohlf2001; @MartinsHansen1997; @OMeara_et_al2006;  @Beaulieu_et_al2012; @Adams2014b; @AdamsCollyer2018b]. These methods are predicated on the notion that phylogenetic signal -- the tendancy for closely related species to display similar trait values -- is present in cross-species datasets [@Felsenstein1985; @Pagel1999; @Blomberg_et_al2003]. Indeed, under numerous evolutionary models, phylogenetic signal is to be expected, as stochastic character change along the hierarchical structure of the tree of life generates trait covaration among related taxa [@Felsenstein1985; @Blomberg_et_al2003; @Revell_et_al2008].

@@

@@

![Response of Pagel's $\lambda$ to increasing strength of Brownian motion. Gray line signifies the 1:1 line where the input value matches the estimate $\hat\lambda$. At each input level, the dark black line represents the empirically derived expected value (mean) of $\hat\lambda$, the red line is the standard deviation of $\hat\lambda$, and the blue line is Shapiro Wilks statistic of $\hat\lambda$ ($W=1.0$ signifies normality, $W< 1.0$ represent skewed distributions).<span data-label="fig:f1"></span>](fig.1.png)

@@

![Response of Blomberg's $\hat\kappa$ to increasing strength of Brownian motion. Gray line signifies the 1:1 line where the input value matches the estimate $\hat\kappa$. At each input level, the dark black line represents the empirically derived expected value (mean) of $\hat\kappa$, the red line is the standard deviation of $\hat\kappa$, and the blue line is Shapiro Wilks statistic of $\hat\kappa$ ($W=1.0$ signifies normality, $W< 1.0$ represent skewed distributions).<span data-label="fig:f2"></span>](fig.2.png)

@@

![Response of effect sizes $Z_{\lambda}$ and $Z_{\kappa}$ to increasing strength of Brownian motion.<span data-label="fig:f3"></span>](fig.3.png)

>>>>

output: rticles::pnas_article

---
Investigating macroevolutionary patterns of trait variation requires a phylogenetic perspective, because the shared ancestry among species violates the assumption of independence among trait values that is common for statistical tests [@Felsenstein1985; @HarveyPagel1991]. Accounting for this evolutionary non-independence is the purview of *phylogenetic comparative methods* (PCMs): a suite of analytical tools that condition trends in the data on the phylogenetic relatedness of observations [@Grafen1989; @GarlandIves2000; @Rohlf2001; @MartinsHansen1997; @OMeara_et_al2006;  @Beaulieu_et_al2012; @Adams2014b; @AdamsCollyer2018b]. These methods are predicated on the notion that phylogenetic signal -- the tendancy for closely related species to display similar trait values -- is present in cross-species datasets [@Felsenstein1985; @Pagel1999; @Blomberg_et_al2003]. Indeed, under numerous evolutionary models, phylogenetic signal is to be expected, as stochastic character change along the hierarchical structure of the tree of life generates trait covaration among related taxa [@Felsenstein1985; @Blomberg_et_al2003; @Revell_et_al2008].

@@

\begin{figure}
\centering
\includegraphics[width=8.25cm]{fig.1.png}
\caption{Response of Pagel's \(\lambda\) to increasing strength of
Brownian motion. Gray line signifies the 1:1 line where the input value
matches the estimate \(\hat\lambda\). At each input level, the dark
black line represents the empirically derived expected value (mean) of
\(\hat\lambda\), the red line is the standard deviation of
\(\hat\lambda\), and the blue line is Shapiro Wilks statistic of
\(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).{}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=8.25cm]{fig.2.png}
\caption{Response of Blomberg's \(\hat\kappa\) to increasing strength of Brownian motion. Gray line signifies the 1:1 line where the input value matches the estimate \(\hat\kappa\). At each input level, the dark black line represents the empirically derived expected value (mean) of
\(\hat\kappa\), the red line is the standard deviation of
\(\hat\kappa\), and the blue line is Shapiro Wilks statistic of
\(\hat\kappa\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).{}}
\end{figure}

@@

@@

\begin{figure}
\centering
\includegraphics[width=8.25cm]{fig.3.png}
\caption{Response of effect sizes \(Z_{\lambda}\) and \(Z_{\kappa}\) to increasing strength of Brownian motion.{}}
\end{figure}
